---
title: "ChiSquare Goodness of fit"
author: "Thuy Tran"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
---

Import the required libraries
```{r}
library(stats)
library(graphics)
library(MASS)
library(dplyr)
```
*In statistics, the chi-squared goodness-of-fit test is used to determine if a categorical dataset fits a theoretical or expected distribution. This test is commonly used when we have observed data in different categories and we want to assess whether the observed frequencies differ significantly from what we would expect under a specified theoretical distribution. The test helps answer questions like "Does the observed data match the expected distribution?"*

The formula:
\[ \chi^2 = \sum \frac{(O - E)^2}{E} \]

Where:
<br> \( \chi^2 \) is the chi-squared statistic.
<br> \( O \) is the observed frequency for each category.
<br> \( E \) is the expected frequency for each category.
<br> The sum is taken over all categories.

**Degrees of Freedom (df):**

\[ df = \text{number of categories} - 1 \]

**The assumptions of the chi-squared test in a more concise form:**

- Independence: Observations must be independent.
- Random Sampling: Data should be collected through random sampling.
- Sample Size: Each expected cell frequency should be ≥ 5.
- Categorical Data: Variables analyzed should be categorical.
- Expected Frequencies: Expected frequencies should be based on reasonable assumptions.
- No Empty Cells: Contingency table should have no empty cells.
- Non-Negative Frequencies: Frequencies in each cell should be non-negative.
**The warning: **
Warning in chisq.test(table(observed_df), p = expected_proportions) :
  Chi-squared approximation may be incorrect
  
To address the chi-squared test warning "Chi-squared approximation may be incorrect," consider these concise strategies:

- Increase Sample Size: Larger samples reduce the risk of small expected frequencies.
- Combine Categories: Merge small categories with neighboring ones, if theoretically justified.
- Fisher's Exact Test: For small samples, use Fisher's Exact Test, providing exact p-values.
- Monte Carlo Simulation: Utilize simulation to handle small expected frequencies and obtain more accurate p-values.

## Problem 1: eye colours
### The dataset
This dataset contains observations of eye colors for 200 individuals including Brown, Blue, Green, and Hazel.

### Hypotheses

<br>H0: The observed distribution of eye colors is consistent with the expected distribution based on the specified proportions.
<br>H1: The observed distribution of eye colors is not consistent with the expected distribution based on the specified proportions.

```{r}
df = read.csv("eye_color.csv")
head(df)
```
**Check all the eye colours**
```{r}
print(unique(df$EyeColor))
```
**Create the frequency table**
```{r}
observed_df<- as.data.frame(table(df$EyeColor))
# Rename the columns to "EyeColor" and "Freq"
colnames(observed_df) <- c("EyeColor", "Freq")
observed_df
```
Now, we want to check whether or not the observed data comply with the expected data:
```{r}
# Notice the order!
expected_proportions <- c(Blue = 0.25, Brown = 0.2, Green = 0.3, Hazel = 0.25)
expected_proportions
```

**Perform the ChiSquare goodness of fit test**

```{r}
chisq_result <- chisq.test(observed_df$Freq, p = expected_proportions)

# Print the chi-squared test result
chisq_result
```
Conclusion: p-value > 0.05 => not enough evidence to reject H0 => accept H0.
In other words, the eye colours comply with the theory.



## Problem 2: Mendel's law of segregation.

### The dataset

A geneticist conducted a cross between two fruit fly individuals, F1, and obtained 176 F2 offspring, consisting of 130 with the wild-type phenotype and 46 with a mutant phenotype. The geneticist wants to test whether the results obtained are consistent with the expected 3:1 ratio predicted by Mendel's Law of Segregation or not. The significance of level is 0.05.

### Hypotheses


<br>H0: The observed ratio of wild-type to mutant offspring in the F2 generation follows the expected 3:1 ratio.
<br>H1: The observed ratio of wild-type to mutant offspring in the F2 generation does not follow the expected 3:1 ratio.

```{r}
# Observed data
expected <- c(130, 46)
# The ratio
ratio <- c(0.75, 0.25)
# Perform the chi-square test
chi_square_test <- chisq.test(expected, p=ratio)

# Print the test result
chi_square_test
```
Conclusion: the p-value > 0.05, there is not enough evidence to reject H0, therefore, accept H0.
